---
title: "Breast Cancer R Analysis Project"
author: Blair Steele
course: BINF 5003 - Data Mining
student ID: n01606240
date: April 7, 2026
output: pdf_document
---

```{r}
# save.image("my_workspace.RData")
# save.image("BcancerRAnalysisProject_workspace.RData")
load("BcancerRAnalysisProject_workspace.RData")

```

----------------------------
RESEARCH QUESTION
----------------------------

Can RNA-seq. multi-gene signatures from TCGA BRCA data be used to predict overall survival in breast cancer patient?

----------------------------
PROJECT PIPLINE
----------------------------

      Data Acquisition
             |
             |
             ▼
Clinical Data Pre-processing & Integration
             |
             |
             ▼
Data Exploration & Visualization
             |
             |
             ▼
Pre-processing & Quality Assessment of RNA-seq Data
             |
             |
             ▼
Principal Component Analysis (PCA)
             |
             |
             ▼
Differential Expression Analysis
             |
             |
             ▼
      Survival Analysis
             |
             |
             ▼
Key Findings & Biological Insights
             |
             |
             ▼
  Conclusion & Future Work

----------------------------

```{r}
library(tidyverse)
library(dplyr)

# install.packages("datasauRus")
library(datasauRus)

suppressPackageStartupMessages({
    library(data.table)
    library(tidyverse)
    library(ggplot2)
    library(ggpubr)
    library(ggsignif)
    library(ggrepel)
})

# install.packages(c("ggpubr", "ggsignif", "ggrepel"))

```

----------------------------
CLINICAL DATA PRE-PROCESSING, TRANSFORMATION, AND INTEGRATION
----------------------------

```{r} 
#Set Working Dictionary
setwd("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2")

file.exists("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2") #TRUE file exists here
read.table("Human__TCGA_BRCA__MS__Clinical__Clinical__01_28_2016__BI__Clinical__Firehose.tsi")

matrix_table <- read.table(
  "Human__TCGA_BRCA__MS__Clinical__Clinical__01_28_2016__BI__Clinical__Firehose.tsi",
  header = TRUE,
  sep = "\t",
  stringsAsFactors = FALSE,
  check.names = FALSE

)

exists("matrix_table") #TRUE
list.files()

dim(matrix_table) #20rows (clinical attributes) 1098columns (# of patients)
# head(matrix_table)
head(matrix_table[, 1:5]) #by default first 6 rows & 5 columns

#TCGA.5L.AAT0 is patient barcode

colnames(matrix_table)
rownames(matrix_table)

col2 <- matrix_table[, 1]
head(col2)  

View(matrix_table)
dim(matrix_table) #20 x 1098 #NOT TIDY FORMAT

#save matrix
# write.csv(matrix_table, "matrix_table.csv", row.names = FALSE)
# matrix_table <- read.csv("matrix_table.csv") #load matrix

``` 

----------------------------
DATA WRANGLING FOR MATRIX TABLE

Matrix_table = 20 rows x 1,098 columns
- As of now, the original clinical matrix is not in tidy format;
      -    The rows = variables such as race, gender, years_to_birth, tumor_purity, pathological_stage
      -    The columns = patient (except thefirst column = attribute names - race, gender, years_to_birth, tumor_purity, pathological_stage etc.)

- Transpose the matrix (i.e., flop the columns and rows), such that
      -    One row = one patient
      -    One column = one variable

```{r}
#TRANSPOSE THE matrix_table

rownames(matrix_table) <- matrix_table [,1] #this makes the attributes into the rows
new_matrix <- matrix_table [, -1] #first column (i.e., attribute names) removed
transposed_matrix <- as.data.frame(t(new_matrix)) #transpose the matrix
transposed_matrix$Patient_Barcode <- rownames(transposed_matrix) #add new column = "patient barcode"
rownames(transposed_matrix) <- NULL

#make patient barcode the second column (i.e, after row number)
cols <- colnames(transposed_matrix)
cols <- cols[cols != "Patient_Barcode"] #removes the column from the position it is now in
cols <- append(cols, "Patient_Barcode", after = 0) #puts column in index 1
transposed_matrix <- transposed_matrix[, cols]

#Change all column names to lowercase.
colnames(transposed_matrix) <- tolower(colnames(transposed_matrix))

#standardize the barcode (remove . and replace with -)
transposed_matrix$patient_barcode <-
      toupper(gsub("\\.", "-", trimws(transposed_matrix$patient_barcode)))

#check for missing values/NAs in patient barcode column in transposed matrix table.
sum(is.na(transposed_matrix$patient_barcode)) #returns 0 N/A values in patient barcode column in transposed matrix table.

#Check for spaces at the end of variables
grep("^\\s|\\s$", colnames(transposed_matrix), value = TRUE) #returns character(0) = means no column names have any leading or trailing spaces.

#fix row numbering after wrangling
rownames(transposed_matrix) <- NULL

# sort transposed_matrix table by patient barcode
# transposed_matrix <- transposed_matrix[
#       order(transposed_matrix$patient_barcode),
# ]

#CHECK FOR TYPOS IN COLUMN NAMES - did a MANUAL check = no typos

#Summary of categorical variables in transposed_matrix
categorical_cols <- sapply(transposed_matrix, is.character) #identifies factor/character columns
print(categorical_cols)
lapply(transposed_matrix[, categorical_cols], unique) #identifies unique values in each categorial column
lapply(transposed_matrix[, categorical_cols], table) #identifies frequency of each unique value in each categorical column
#notice "NAs"

#CHECK FOR TYPOS/INCONSISTENT CODING IN CATEGORICAL DATA (race, gender, pathogical stage)
#Inconsistent coding may include entries such as "Male", "male", "M", "Female", "F", "Femlae" etc.

#CHECKING FOR MISSINGNESS (i.e., missing or N/A values)
sum(is.na(transposed_matrix)) #4144 NAs 

#droplevels? to remove NAs? 
#what columns have N/A

#Change format from TCGA.5L.AAT0 to TCGA-5L-AAT0 (like in the new aggregated patient data set)
transposed_matrix$patient_barcode <- toupper(gsub("\\.", "-", trimws(transposed_matrix$patient_barcode))) #now the patient barcodes are in the same format in both tables

# dim(transposed_matrix) #1097 x 21
View(transposed_matrix) #new 'transposed clinical matrix' table after wrangling

#save matrix
# write.csv(transposed_matrix, "transposed_matrix.csv", row.names = FALSE)
# transposed_matrix <- read.csv("transposed_matrix.csv") #load matrix

```

----------------------------
COMBINE ALL TSV FILES 
("clinical.tsv", "exposure.tsv", "family_history.tsv", "follow_up.tsv" and "pathology_detail.tsv")

```{r} 
library(dplyr)
library(readr)

setwd("/Users/BGS/Desktop/🧬 CLINICAL BIOINFORMATICS - HUMBER (LU = Dec.4.2025)/★ SCHOOL/★ SEMESTER 1/BINF 5003 - Data Mining, Modeling and Biostatistics/ASSIGNMENTS ☆ Data Mining, Modeling and Biostatistics - BINF 5003/RNA Analysis Assignment - Breast Cancer 2/clinical.project-tcga-brca.2026-02-12")

files <- list.files(pattern = "\\.tsv$") #lists all tsv files in the folder
print(files)
# View(files)

#READ .tsv FILES
all_tsv_files <- lapply(files, function(f) {
      all_tsv_df <- read_tsv(f, show_col_types = FALSE)
      colnames(all_tsv_df) <- trimws(colnames(all_tsv_df)) #trimws removes white spaces

      if ("cases.submitter_id" %in% colnames(all_tsv_df)) {
        all_tsv_df <- all_tsv_df %>%
            filter(!is.na(cases.submitter_id) & cases.submitter_id != "") %>% #removes missing NA values. "cases.submitter_id" is essentially the patient barcode (rename later)
            group_by(cases.submitter_id) %>%
            slice(1) %>% #ensures one row per patient
            ungroup()
      }
      all_tsv_df 

})

#MERGE ALL TSV FILES BY "cases.submitter_id"
combined_patient_tsv_files <- Reduce(function(x, y) {
      full_join(x, y, by = "cases.submitter_id")
}, all_tsv_files)

#ensure duplicated columns, if any, are removed
combined_patient_tsv_files <-
      combined_patient_tsv_files[, !duplicated(colnames(combined_patient_tsv_files))]

#rename "cases.submitter_id" to "patient_barcode"
colnames(combined_patient_tsv_files)[colnames(combined_patient_tsv_files) == "cases.submitter_id"] <- "patient_barcode"

#check if any NAs are in the patient barcode column
any(is.na(combined_patient_tsv_files$patient_barcode)) #FALSE - no NA's

#move "patient_barcode" to index 1
cols <- colnames(combined_patient_tsv_files)
cols <- cols[cols != "patient_barcode"]
cols <- c("patient_barcode", cols)
combined_patient_tsv_files <- combined_patient_tsv_files[, cols]

#check for spaces in column names 
grep("^\\s|\\s$", colnames(combined_patient_tsv_files), value = TRUE) #returns character(0) = means no column names have any leading or trailing spaces.

#note = #column names are already lowercase (common practice)

# head(combined_patient_tsv_files)
View(combined_patient_tsv_files)
# colnames(combined_patient_tsv_files)
# dim(combined_patient_tsv_files) #1098 x 543

# NOTE = the combined_patient_tsv_files is now in tidy format (row = patient, column = variables), and does not need to be transposed. 

#save matrix
# write.csv(combined_patient_tsv_files, "combined_patient_tsv_files.csv", row.names = FALSE)
# combined_patient_tsv_files <- read.csv("combined_patient_tsv_files.csv") #load matrix

```

```{r}
#Number of males and females in study 
table(combined_patient_tsv_files$demographic.gender) #Males = 12, Females = 1085

#check for inconsistent coding entries such as "Male", "male", "M", "Female", "F", "Femlae" etc.
unique(combined_patient_tsv_files$demographic.gender) #only "female", "male", and "'--"  exit

```

----------------------------
MERGE BOTH TRANSPOSED MATRIX (CLINICAL DATA) & COMBINED PATIENT TSV FILE MATRIX

# CHECK FOR TYPOS IN COLUMN NAMES - manual check
# CHECK FOR TYPOS/INCONSISTENT CODING IN CATEGORICAL DATA (race, gender, pathogical stage)

```{r}
library(dplyr)

clinical_cols <- setdiff(colnames(transposed_matrix), "patient_barcode")  #identifies all columns of the transposed matrix except the "patient_barcode" column

new_clinical_cols <- clinical_cols[!clinical_cols %in% colnames(combined_patient_tsv_files)] #ensures that only unique clinical data from the transposed matrix is added to the combined_patient_tsv_files matrix

transposed_subset <- transposed_matrix %>%
      select(patient_barcode, all_of(new_clinical_cols)) #create a tidy subset of all unique variables + the patient barcodes

final_aggregated_patient_data <- combined_patient_tsv_files %>%
      left_join(transposed_subset, by = "patient_barcode") #matches variables to patient barodes

all_cols <- c("patient_barcode", new_clinical_cols,
      setdiff(colnames(final_aggregated_patient_data), c("patient_barcode", new_clinical_cols)))

final_aggregated_patient_data <- final_aggregated_patient_data[, all_cols] #reorders the columns

# dim(final_aggregated_patient_data) #1098 x 536
View(final_aggregated_patient_data)

#save matrix
# write.csv(final_aggregated_patient_data, "final_aggregated_patient_data.csv", row.names = FALSE)
# final_aggregated_patient_data <- read.csv("final_aggregated_patient_data.csv") #load matrix

```

----------------------------
EXPLORATORY ANALYSIS - VISUALIZATION OF THE RELATIONSHIPS BETWEEN IMPORTANT CLINICAL VARIABLES

```{r} 
#NOTE
#CATEGORICAL variables = gender, race, status, ethnicity, pathologic stage, pathology_T_stage, pathology_N_stage, pathology_M_stage, radiation therapy, HER2.Status, PR.Status, ER.Status, PAM50, histological type
#CATEGORICAL variables should be a FACTOR data type.
#for categorical variables, convert from CHARACTER to FACTOR.

#CONTINUOUS variables such as age (i.e., years to birth), tumor purity, overallsurvival, overall_survival, median overall survival, number of lymph nodes
#CONTINUOUS variables have to be NUMERIC.
#convert from CHARACTER to NUMERIC. 

#remove N/As
#Warning message: NAs introduced by conversion?

```

```{r} 
#AGE DISTRIBUTION

#convert to numeric
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])

#remove any N/A's 
age <-age[!is.na(age)]

#AGE DISTRIBUTION HISTOGRAM
hist(age,
     main = "Distribution of Age",
     xlab = "Age (In Years)",
     col = "yellowgreen",
     breaks = 20)

#save object
# png("age_histogram.png", width = 800, height = 800)
# plot(x, y)
# dev.off()

#youngest and oldest patients
age <- as.numeric(transposed_matrix$years_to_birth)
age <- age[!is.na(age)] #remove NAs 
min(age) #26
max(age) #90
mean(age) #58.59
median(age) #59
sd(age) #13.19

# Takeaway = youngest patient = 26, oldest patient = 90
```

```{r} 
#TUMOR PURITY DISTRIBUTION

#convert to numeric
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"])

#remove any N/A's 
purity <-purity[!is.na(purity)]

#TUMOR PURITY HISTOGRAM
hist(purity, 
    main = "Tumor Purity Distribution", 
    xlab = "Purity", 
    col="#e69e41", 
    breaks =20)

#takeaway = high tumor purity = more cancer cells than normal cells in the tissue.
```

```{r} 
#TUMOR PURITY vs. AGE

#convert to numeric
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"])

purity_age_df <- data.frame(Age = age, TumorPurity = purity)
purity_age_df <- purity_age_df[complete.cases(purity_age_df), ] #this removes the rows that have an NA

#TUMOR PURITY vs. AGE PLOT
plot(purity_age_df$Age, purity_age_df$TumorPurity,
      xlab = "Age (Years)",
      ylab = "Tumor Purity",
      main = "Tumor Purity v. Age",
      pch =16, #plot point = solid circle
      col = "red") 

#takeaway = no apparent trend. Tumor purity does not seem to correlate with age.
# What do the outliers mean?

```

```{r} 
#TUMOR PURITY vs. OVERALL SURVIVAL

#convert to numeric
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"]) #extracts tumor_purity column
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

#create dataframe
df_surv <- data.frame(Purity = purity, OverallSurvival = overall_survival)

#remove N/A's
df_surv <- df_surv[!is.na(df_surv$Purity) & !is.na(df_surv$OverallSurvival), ]

plot(df_surv$Purity, df_surv$OverallSurvival,
    xlab = "Tumor Purity",
    ylab = "Overall Survival (days)",
    main = "Tumor Purity v. Overall Survival",
    pch = 16, col = "#a9db2b")

#Linear Trend Line
abline(lm(OverallSurvival ~ Purity, data = df_surv), col= "#ff5500", lwd=2)

#takeaway = 
#Question: Do tumors with lower purity (more stromal/immune infiltration) correlate with survival?
```

```{r} 
#TUMOR PURITY v. STAGE

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
purity <- as.numeric(transposed_matrix[ ,"tumor_purity"]) #convert to numeric

#create data.frame
df_purity_stage <- data.frame(Stage=stage, Purity=purity)

#remove N/A's
df_purity_stage <- df_purity_stage[!is.na(df_purity_stage$Stage) & !is.na(df_purity_stage$Purity), ]

boxplot(Purity ~ Stage, data=df_purity_stage, 
      main="Tumor Purity v. Stage", 
      col="powderblue")

#takeaway = tumor purity ~0.8 in all stages?

```

EXPLORATION OF PAM50 SUBTYPES
```{r}
table(transposed_matrix$pam50)

# Basal  Her2  LumA  LumB 
#   147    67   426   186 

#Most patients have the Luminal A molecular subtype of breast cancer.
```

```{r} 
#TUMOR PURITY v. PAM50

# pam50 <- unlist(transposed_matrix[ ,"PAM50"]) #extracts PAM50 column
# purity <- as.numeric(transposed_matrix[ ,"tumor_purity"]) #convert to numeric

# #create dataframe
# df_pam_50_new <- data.frame(PAM50Subtype=pam50, Purity=purity)

# #remove N/As 
# df_pam_50_new <- df_pam_50_new[!is.na(df_pam_50_new$PAM50Subtype) & !is.na(df_pam_50_new$Purity), ]

# boxplot(Purity ~ PAM50Subtype, data=df_pam_50_new, 
#       main="Tumor Purity v. PAM50", 
#       xlab = "PAM50 Subtype",
#       ylab = "Tumor Purity",
#       col="#b435b6")

#PAM50 is a measure of gene expression profile that allows for the classification of 
#Breast cancer into four biologically distinct molecular subtypes - Luminal A, Luminal B, HER2-Enriched, Basal-Like, Normal-Like

#takeaway = which molecular subtype has the highest/lowest tumor purity? is this consistent with facts?

```

```{r} 
#SURVIVAL DISTRIBUTION

#remove N/As
overall_survival <- overall_survival[!is.na(overall_survival)]

#Plot Histogram
hist(overall_survival, 
      main = "Overall Survival Distribution", 
      xlab="Survival (days)", 
      col= "tomato2", 
      breaks=20)

#takeaway = most people died w/in 2000 days of diagnosis?
```

```{r} 
#AGE vs. SURVIVAL OUTCOME

#convert to numeric
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])
age <- as.numeric(transposed_matrix[ ,"years_to_birth"])

#create dataframe
df <- data.frame(Age = age, OverallSurvival = overall_survival)

#Remove N/As
df <- df[complete.cases(df), ]

plot(df$Age, df$OverallSurvival,
    xlab = "Age",
    ylab = "Overall Survival (Days)",
    main = "Patient Age v. Survival Outcome",
    col = "#dc9107",
    pch = 16) 

#Trendline
abline(lm(OverallSurvival ~ Age, data=df), col="blue", lwd=2)

#takeaway = age is inversely proportion of survival outcome. The older the patient, the lower the survival outcome.
```

```{r} 
#OVERALL SURVIVAL BY STAGE

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

#create dataframe 
df2 <- data.frame(Stage = stage, Survival = overall_survival)

boxplot(Survival ~ Stage, data = df2,
        main = "Overall Survival by Stage",
        col = "mistyrose")

#takeaway = overall survival decreases by stage (stage IV = lesser days of survival)
```

```{r} 
#NUMBER OF LYMPH NODES v. OVERALL SURVIVAL

lymph_nodes <- as.numeric(transposed_matrix[ , "number_of_lymph_nodes"])
overall_survival <- as.numeric(transposed_matrix[ , "overall_survival"])

#create dataframe
df_ln <- data.frame(LymphNodes = lymph_nodes, OverallSurvival = overall_survival)

#remove N/As
df_ln <- df_ln[complete.cases(df_ln), ]

plot(df_ln$LymphNodes, df_ln$OverallSurvival,
    xlab = "Number of Lymph Nodes",
    ylab = "Overall Survival (Days)",
    main = "Number of Lymph Nodes v. Overall Survival",
    col = "#29aaef",
    pch = 16)

#takeaway = lesser lymph node involvement = longer survival. more lymph node involvement = more cancer activity (i.e., sign of spread)

```

```{r} 
#PATHOLOGIC STAGE DISTRIBUTION

stage <- unlist(transposed_matrix[ ,"pathologic_stage"]) #extracts pathological_stage column
barplot(table(stage), main="Pathologic Stage Distribution", col="#039a9a", las=2)

#takeaway = most patients are in stage 2. stage 4 has the least number of patients.
```

```{r} 
# CORRELATION CHECK BETWEEN CONTINUOUS VARIABLES:

#Age v. Tumor Purity 
cor(age, purity, use="complete.obs") #0.082
#Takeaway = weak POSITIVE correlation between AGE & PURITY.
#As age increases, purity only slightly increases.
#Close to 0 = almost no relationship.

#Age v. Overall Survival
cor(age, overall_survival, use="complete.obs") #-0.168
#Takeaway = weak NEGATIVE correlation between AGE & SURVIVAL.
#Age Increases, Overall Survival Decreases, but weak relationship.

#Tumor Purity v. Overall Survival
cor(purity, overall_survival, use="complete.obs") #-0.007
#Takeaway = weak NEGATIVE correlation between PURITY & SURVIVAL. 
#Small NEGATIVE correlation = suggests tumor purity and overall survival are independent of each other. 

```

----------------------------
PRE-PROCESSING & QUALITY ASSESSMENT OF RNA-SEQ GENE EXPRESSION DATA.

```{r} 
# getwd()

#LOAD RNA SEQUENCE DATA (.cct file)
lines <- readLines ("Human__TCGA_BRCA__UNC__RNAseq__GA_RNA__01_28_2016__BI__Gene__Firehose_RSEM_log2.cct", n = 5)
print(lines)

RNA_seq_data <- read.table("Human__TCGA_BRCA__UNC__RNAseq__GA_RNA__01_28_2016__BI__Gene__Firehose_RSEM_log2.cct", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

#change "gene name" to "gene_name"
RNA_seq_data <- RNA_seq_data %>%
      rename('gene_name' = 'attrib_name')


#fix format of patient barcodes from TCGA.A1.A0SD to TCGA-A1-A0SD (consistent across all datasets)
colnames(RNA_seq_data)[-1] <- gsub("\\.", "-", colnames(RNA_seq_data)[-1])

# colnames(RNA_seq_data)
# str(RNA_seq_data)
# head(RNA_seq_data)
# nrow(RNA_seq_data) #6149 rows = 6149 genes
View(RNA_seq_data)
# summary(RNA_seq_data)

## RNA_seq_table is just RNA_seq_data without the gene_name column

#table contains mixed data types
#convert all columns except attrib_name to NUMERIC. 
RNA_seq_table <- RNA_seq_data[, -1] #removes gene name column (i.e., the column with the names of the genes). 
RNA_seq_table <- data.frame(lapply(RNA_seq_table, function(x) as.numeric(as.character(x)))) #converts to numeric
RNA_seq_table <- as.matrix(RNA_seq_table) #converts to matrix

#convert to matrix???
RNA_seq_table <- as.matrix(RNA_seq_table)
class(RNA_seq_table) #check that it has been converted to a dataframe

#standardize column names (replace . for - in patient barcodes)
colnames(RNA_seq_table) <- toupper(
      gsub("\\.", "-", trimws(colnames(RNA_seq_table)))
)

View(RNA_seq_table)

#save matrix as csv
# write.csv(RNA_seq_data, "RNA_seq_data.csv", row.names = FALSE)
# RNA_seq_data <- read.csv("RNA_seq_data.csv") #load matrix

# write.csv(RNA_seq_table, "RNA_seq_table.csv", row.names = FALSE)
# RNA_seq_table <- read.csv("RNA_seq_table.csv") #load matrix

```

```{r}
library(dplyr)
library(tidyr)

#CONVERT TIDY FORMAT: #3,233,933 rows?????
# each row should have one observation (i.e., one gene per row)
# each column should have one variable

RNA_seq_tidy <- RNA_seq_data %>%
      pivot_longer(
      cols = -gene_name, #preserves the gene name
      names_to = "sample",
      values_to = "expression"
      ) %>%
      mutate(expression = as.numeric(expression))

#REMOVE ROWS WITH N/A
RNA_seq_tidy <- na.omit(RNA_seq_tidy)

#convert column names to lower case (best practice)
colnames(RNA_seq_tidy) <- tolower(colnames(RNA_seq_tidy))

#rename "sample" column to "Patient Barcode"
RNA_seq_tidy <- RNA_seq_tidy %>%
      rename("patient_barcode" = sample)

#move 'patient_barcode' column to index 1
RNA_seq_tidy <- RNA_seq_tidy %>%
      select("patient_barcode", everything())

#fixes format of patient barcodes from TCGA.A1.A0SD to TCGA-A1-A0SD (consistent across all datasets)
RNA_seq_tidy$patient_barcode <- gsub("\\.", "-", RNA_seq_tidy$patient_barcode)

# colnames(RNA_seq_tidy)
# #change "gene name" to "gene_name"
# RNA_seq_tidy <- RNA_seq_tidy %>%
#       rename("gene_name" = 'attrib_name')

#change "expression" to "expression_log2rsem"
RNA_seq_tidy <- RNA_seq_tidy %>%
      rename("expression_log2rsem" = 'expression')

# colnames(RNA_seq_tidy)
# nrow(RNA_seq_tidy) #3,233,933 rows?????
# head(RNA_seq_tidy)
# View(RNA_seq_tidy)

# RNA-seq now in long/tidy format, i.e., :
# - row = patient
# - columns = gene expression variables (or gene + expression)

#save matrix as csv
# write.csv(RNA_seq_tidy, "RNA_seq_tidy.csv", row.names = FALSE)
# # RNA_seq_tidy <- read.csv("RNA_seq_tidy.csv") #load matrix

```

HISTOGRAM OF LOG2-TRANSFORMED GENE EXPRESSION
```{r}
library(ggplot2)

log2_histogram <- ggplot(RNA_seq_tidy, aes(x = expression)) +
                        geom_histogram(bins = 50, fill = "steelblue", color = "black") +
                        theme_minimal() +
                        labs(
                              title = "Distribution of Log2-Transformed Gene Expression",
                              x = "Log2 Expression (RSEM)",
                              y = "Frequency"
                        )

print(log2_histogram)

# ggsave(
#       filename = "log2_histogram.png",
#       plot = log2_histogram,
#       width = 8,
#       height = 6,
#       dpi = 300
# )

```

IDENTIFICATION OF HIGHLY EXPRESSED GENES (NOT NECESSARILY CANCER RELATED)

```{r} 
#IDENTIFICATION OF BREAST CANCER GENES

b_cancer_genes <- (RNA_seq_tidy$gene_name)
# print(b_cancer_genes) #prints only column 1 of RNA Seq Data (i.e., the genes involved in BCancer)
length(b_cancer_genes) 

unique_genes <- unique(RNA_seq_tidy$gene_name) #identification of unique genes
length(unique_genes) #6149 UNIQUE genes
# print(unique_genes) #they're all unique

#save matrix as csv
# write.csv(unique_genes, "unique_genes.csv", row.names = FALSE)
# unique_genes <- read.csv("unique_genes.csv") #load matrix
```

MEAN GENE EXPRESSION ANALYSIS AND GENE RANKING

```{r}
library(dplyr)

#Calculate mean expression per gene
mean_gene_expression_dframe <- RNA_seq_data %>%
      rowwise() %>% 
      mutate(mean_expression = mean(c_across(-gene_name), na.rm = TRUE)) %>%
      ungroup() %>%
      select(gene_name, mean_expression) 

##TOP 20 HIGHEST EXPRESSED BREAST CANCER GENES
top_20_exp_genes <- mean_gene_expression_dframe %>%
      arrange(desc(mean_expression)) %>% #sort gene expression means in decreasing order

      slice_head(n = 20)

View(top_20_exp_genes)
# print(top_20_exp_genes)

#save matrix as csv
# write.csv(top_20_exp_genes, "top_20_exp_genes.csv", row.names = FALSE)
# top_20_exp_genes <- read.csv("top_20_exp_genes.csv") #load matrix

```

VISUALIZATION & INTERPRETATION OF HIGHLY EXPRESSED GENES

```{r}
#BARPLOT VISUALIZATION OF THE TOP 20 HIGHLY EXPRESSED GENES

library(ggplot2)
library(dplyr)

#create ggplot barplot
top20_exp_plot <- ggplot(top_20_exp_genes, aes(x = reorder(gene_name, mean_expression), y = mean_expression)) +
            geom_bar(stat = "identity", fill = "limegreen") +
            coord_flip() + #rearrange the coordinates so that we can easily read the gene names
            labs(
                  title = "Top 20 Highest Expressed Breast Cancer Genes",
                  x = "Genes",
                  y = "Mean Gene Expression (Log2 RSEM)"
            ) +
            theme_minimal() +
            theme(
                  axis.text.x = element_text(angle = 0, hjust = 0.5), #x-axis labels
                  axis.text.y = element_text(size = 10),              #y-axis labels (gene names)
                  axis.title.x = element_text(size = 12, face = "bold"),
                  axis.title.y = element_text(size = 12, face = "bold"),
                  plot.title = element_text(size = 14, face = "bold", hjust = 0.5) 
      )

print(top20_exp_plot)

#take away = the following genes are the most highly expressed genes across patients
#ASPN, RGS1, SCGB2A2, MMP7, COL10A1, FNDC1, CD93, ADAMDEC1, IGSF6, HLA-DRA, SCGB1D2, CPXM2, KLRB1, F13A1, C10orf81, COMP, CXCL13, MYH11, IRF6, CD34
#but these could be housekeeping- and other types of genes that do not have any prognostic significant.
#next is the ID of survival-associated gene

#save object as PNG
# ggsave("Top_20_Highest_Expressed_Breast_Cancer_Genes.png", plot = top20_exp_plot, width = 6, height = 6, dpi = 300)

#save as R object
# save(top20_exp_plot, file = "top20_exp_plot.RData")
# load("top20_exp_plot.RData")

```

ALTHOUGH I HAVE IDENTIFIED THE TOP 20 HIGHEST EXPRESSED GENES IN THIS BCANCER STUDY, 
IT DOESN'T NECESSARILY MEAN THAT THESE TOP 20 GENES HAVE BIOLOGICAL OR PROGNOSTIC IMPORTANCE (I.E., THEY ARE NOT NECESSARILY SURVIVAL-ASSOCIATED GENES).
THEY COULD BE HOUSEKEEPING GENES, IMMUNE FILTRATION GENES, OR STROMAL GENES THAT ARE 
HIGHLY EXPRESSED ACROSS ALL TUMORS. 

THE NEXT STEP AFTER THIS EXPLORATORY ANALYSIS STEP IS TO PREP. THE DATA FOR PCA & limma.
----------------------------

DATA PREPARATION FOR DOWNSTREAM MODELING. 

```{r}

lim_matrix <- as.matrix(RNA_seq_data[, -1])
rownames(lim_matrix) <- RNA_seq_data$gene_name

#expression matrix
lim_matrix <- as.matrix(RNA_seq_data[, -1])
rownames(lim_matrix) <- RNA_seq_data$gene_name

pam50 <- factor(final_aggregated_patient_data$pam50) #subset the PAM50 column from final_aggregated_patient_data and convert to factor (so that R knows we are dealing with categorical data, as opposed to continuous data).
names(pam50) <- final_aggregated_patient_data$patient_barcode #assign PAM50 labels to respective patient barcodes (LumA, LumB, HER2, Basal)

#remove NAs from PAM50 patients 
pam50 <- pam50[!is.na(pam50)]

#match PAM50 samples with respective RNA seq data via patient barcode
common_samples <- intersect(colnames(lim_matrix), names(pam50)) #common samples = patients that have both PAM50 labels and RNA_seq data
lim_matrix <- lim_matrix[, common_samples]
pam50 <- pam50[common_samples]
pam50 <- droplevels(pam50) #removes unused subtypes from the data. 

# ncol(lim_matrix)  #445 - not all patients have both PAM50 data and RNA_seq data. 
View(lim_matrix)

#align
all(colnames(lim_matrix) == names(pam50))

#design matrix
design <- model.matrix(~0 + pam50)
colnames(design) <- levels(pam50)
rownames(design) <- names (pam50) #add patient barcode to design matrix

View(design)
# nrow(design) #445 - not all patients have both PAM50 data and RNA_seq data.
#remember for limma, ncol(lim_matrix) == nrow(design)

```
----------------------------

PRINCIPAL COMPONENT ANALYSIS(PCA) ON RNA-SEQ DATA
PCA REDUCES HIGH DIMENSIONALITY DATA. 

```{r}
gene_variance <- apply(lim_matrix, 1, var)

top_genes <- names(sort(gene_variance, decreasing = TRUE))[1:500] #select the top 500 most variable genes

pca_input_top <- t(lim_matrix[top_genes, ]) #transpose matrix so rows = samples and columns = genes

pca_top <- prcomp(pca_input_top, scale. = TRUE)

pca_top_df <- as.data.frame(pca_top$x) #converts output of PCA to a dataframe

pca_top_df$PAM50 <- pam50 #adds pam50 labels

#PCA Plot of Top 100 Variable Genes
pca_plot <- ggplot(pca_top_df, aes(PC1, PC2, color = PAM50)) +
                  geom_point(size = 2) +
                  theme_minimal() +
                  labs(title = "PCA of Top 500 Variable Genes")

print(pca_plot)

#save ggplot
# ggsave("PCA Plot.png", plot = pca_plot, width = 8, height = 6, dpi = 300)

```

Interpretation of PCA plot:

PCA plot show how the samples cluster based on the top 100 most variable genes.
Colours represent the PAM50 subtypes.
      - Strong separation of Basal subtype (red) from the other subtypes = indicates that Basal tumors have a distinct gene expression profile. Low PC1 values (i.e., tight clustering on the left)
      - LumA (teal) and LumB (purple) show substantial overlapping = indicates biological similarity. Higher PCA (clustering on the right side).
      - HER2 (green) Less tightly clustered. Spread out between Basal and Luminal groups = suggests greater heterogeneity or mixed expression patterns.

PC1 is the largest source of variation (by nature of PCA)

----------------------------

SCREE PLOT 

```{r}
var_explained <- pca_top$sdev^2 / sum(pca_top$sdev^2) #explanation of the variance

#scree dataframe
scree_df <- data.frame(
      PC = 1:10, #create dataframe of first 10 principal components.
      Variance = var_explained[1:10]
)

#scree plot
scree_plot <- ggplot(scree_df, aes(x = PC, y = Variance)) +
      geom_line() +
      geom_point() +
      labs(
            title = "Scree Plot",
            x = "Principal Component",
            y = "Variance Explained"
      ) +
      theme_minimal()

print(scree_plot)

#save ggplot
# ggsave("Scree Plot.png", plot = scree_plot, width = 8, height = 6, dpi = 300)

```

Interpretation of Scree plot:

      - PC1 = captures the largest source of variation = suggests biological separation of subtypes.
      - PC2 & PC3 = captures smaller sources of variation (likely differences within Luminal and HER2 subtypes)

----------------------------
----------------------------
DIFFERENTIAL GENE EXPRESSION ANALYSIS WITH LIMMA

Afer ID of these significant genes, Cox regression (survival analysis) will be conducted
on said subset of genes (i.e., only the biologically significant genes)

LIMMA Differential Analysis is the foundation for:
      - LASSO-Cox
      - Predictive modeling
      - Gene signature construction

LIMMA can be used on both continuous and categorical data

```{r}
library(limma)
library(tibble)

#fit the linear model
fit <- lmFit(lim_matrix, design) #lmFit performs a linear regression on each of the 6149 genes to find differences in gene expression across PAM50 subtypes. 
fit <- eBayes(fit)

#save design_matrix as csv
# write.csv(design, "design_matrix.csv", row.names = TRUE)
# design <- read.csv("design_matrix.csv") #load matrix

#save lim_matrix as csv
# write.csv(lim_matrix, "lim_matrix.csv", row.names = TRUE)
# lim_matrix <- read.csv("lim_matrix.csv") #load matrix

```

SUBTYPE-SPECIFIC DIFFERENTIAL EXPRESSION

```{r}
library(limma)

head(fit$coefficients) #gene expression estimates per subtype

diff_exp_results <- topTable(fit, number = Inf, adjust.method = "BH") #creates a ranked table of the genes that are most likely to be differentially expressed. #BH = Benjamini-Hochberg procedure = controls false discovery rate (i.e., the proportion of false positives among significant genes)

diff_exp_results <- rownames_to_column(diff_exp_results, var = "gene_name") #convert rownames to a column called gene names

# head(diff_exp_results)
View(diff_exp_results)

# Differentially Expressed Genes (FDF < 0.05)
sum(diff_exp_results$adj.P.Val < 0.05) #6047 #suggests that all but 102 genes are differentially expressed. 
#see next code block for calculation of genes differentially expressed across subtype contrasts (FDF < 0.05)

# 1,565-4,348 significantly different genes per subtype comparison (FDR <0.05). 

#save subtype specific differential expression results as csv
# write.csv(diff_exp_results, "diff_exp_results.csv", row.names = TRUE)
# diff_exp_results <- read.csv("diff_exp_results.csv") #load matrix

```

PAIR-WISE COMPARISONS OF GENE EXPRESSION.

      - Basal_vs_LumA
      - Basal_vs_LumB
      - Basal_vs_Her2
      - LumA_vs_LumB
      - LumA_vs_Her2 *
      - LumB_vs_Her2

```{r}
#pair-wise PAM50 subtype contrasts 
contrast.matrix <- makeContrasts(
      LumA_vs_Basal = LumA - Basal,
      LumB_vs_Basal = LumB - Basal,
      Her2_vs_Basal = Her2 - Basal,
      LumA_vs_LumB = LumA - LumB,
      LumA_vs_Her2 = LumA - Her2,
      LumB_vs_Her2 = LumB - Her2,
      levels = design
)

# contrast.matrix <- makeContrasts(
#       LumA_vs_Basal = pam50LumA - pam50Basal,
#       LumB_vs_Basal = pam50LumB - pam50Basal,
#       Her2_vs_Basal = pam50Her2 - pam50Basal,
#       LumA_vs_LumB = pam50LumA - pam50LumB,
#       LumA_vs_Her2 = pam50LumA - pam50Her2,
#       LumB_vs_Her2 = pam50LumB - pam50Her2,
#       levels = design
# )

fit2 <- contrasts.fit(fit, contrast.matrix) #apply contrasts
fit2 <- eBayes(fit2) #stabilizes gene variance estimates across all genes, thus improving statistical power = more reliable p-values.

#F-tests
diff_exp_pair_results <- topTableF(fit2, number = Inf) #shows the total # of genes differentially expressed across all subtypes. topTableF performs ANOVA-style F-test across all PAM50 subtypes. 
sum(diff_exp_pair_results$adj.P.Val < 0.05) #5100 of 6149 genes total.

head(diff_exp_pair_results)
View(diff_exp_pair_results)

#shows the total # of genes differentially expressed across subtype contrasts.
LumA_vs_Basal_results <- topTable(fit2, coef = "LumA_vs_Basal", number = Inf)
LumB_vs_Basal_results <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)
Her2_vs_Basal_results <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)
LumA_vs_LumB_results <- topTable(fit2, coef = "LumA_vs_LumB", number = Inf)
LumA_vs_Her2_results <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)
LumB_vs_Her2_results <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

#calculation of genes differentially expressed across subtype contrasts (FDF < 0.05)
sum(LumA_vs_Basal_results$adj.P.Val < 0.05) #4348 genes differentially expressed across LumA_vs_Basal
sum(LumB_vs_Basal_results$adj.P.Val < 0.05) #3740 genes differentially expressed across LumB_vs_Basal
sum(Her2_vs_Basal_results$adj.P.Val < 0.05) #2609 genes differentially expressed across Basal_vs_Her2
sum(LumA_vs_LumB_results$adj.P.Val < 0.05) #2687 genes differentially expressed across LumA_vs_LumB
sum(LumA_vs_Her2_results$adj.P.Val < 0.05) #2564 genes differentially expressed across LumA_vs_Her2
sum(LumB_vs_Her2_results$adj.P.Val < 0.05) #1565 genes differentially expressed across LumB_vs_Her2

#save subtype specific differential expression pair-wise results as csv
# write.csv(diff_exp_pair_results, "diff_exp_pair_results.csv", row.names = TRUE)
# diff_exp_pair_results <- read.csv("diff_exp_pair_results.csv") #load matrix

```

----------------------------
----------------------------
PAIR-WISE CONTRAST MATRICES & VOLCANO PLOTS

LumA v. Basal Contrast Matrix
```{r}
lumA_basal <- topTable(fit2, coef = "LumA_vs_Basal", number = Inf)

#Add gene name to lumA_basal matrix
lumA_basal <- lumA_basal %>% rownames_to_column(var = "gene_name")

View(lumA_basal)

#Interpretation
#positive logFC (e.g., logFC = 2), means the gene is expressed 4x more in LumA tumors than Basal tumors 
#negative logFC (e.g., logFC = -2), means gene is express 4x more in Basal tumors than LumA tumors

#save lumA_basal_matrix as csv
# write.csv(lumA_basal, "lumA_basal_matrix.csv", row.names = TRUE)
# lumA_basal <- read.csv("lumA_basal_matrix.csv") #load matrix
```

LumA v. Her2 Contrast Matrix
```{r}
lumA_Her2 <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)

#Add gene name to lumA_Her2 matrix
lumA_Her2 <- lumA_Her2 %>% rownames_to_column(var = "gene_name")

View(lumA_Her2)

#Interpretation
#positive logFC = higher gene expression in LumA. 
#negative logFC = higher gene expression in Her2.

#save lumA_her2_matrix as csv
# write.csv(lumA_Her2, "lumA_her2_matrix.csv", row.names = TRUE)
# lumA_Her2 <- read.csv("lumA_her2_matrix.csv") #load matrix
```

LumB v. Basal Contrast Matrix
```{r}
lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)

#Add gene name to lumB_basal matrix
lumB_basal <- lumB_basal %>% rownames_to_column(var = "gene_name")

View(lumB_basal)

#Interpretation
#positive logFC = higher gene expression in LumB.
#negative logFC = higher gene expression in Basal.

#save lumA_her2_matrix as csv
# write.csv(lumB_basal, "lumB_basal_matrix.csv", row.names = TRUE)
# lumB_basal <- read.csv("lumB_basal_matrix.csv") #load matrix

```

Her2 vs. Basal Contrast Matrix
```{r}
Her2_basal <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)

#Add gene name to Her2_basal matrix
Her2_basal <- Her2_basal %>% rownames_to_column(var = "gene_name")

View(Her2_basal)

#Interpretation
#positive logFC = higher gene expression in Her2.
#negative logFC = higher gene expression in Basal.

#save Her2_Basal_matrix as csv
# write.csv(Her2_basal, "Her2_vs_Basal_matrix.csv", row.names = TRUE)
# Her2_vs_Basal <- read.csv("Her2_vs_Basal_matrix.csv") #load matrix

```

LumB v. Her2 Contrast Matrix
```{r}
#lumB_basal contrast
lumB_Her2 <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

# #Add gene name to lumA_Her2 matrix
lumB_Her2 <- lumB_Her2 %>% rownames_to_column(var = "gene_name")

View(lumB_Her2)

#Interpretation
#positive logFC = higher gene expression in lumB.
#negative logFC = higher gene expression in Her2.

#save lumA_her2_matrix as csv
# write.csv(lumB_Her2, "lumB_Her2.csv", row.names = TRUE)
# lumB_Her2 <- read.csv("lumB_Her2_matrix.csv") #load matrix

```

----------------------------
----------------------------
DIFFERENTIAL EXPRESSION FILTERING & IDENTIFICATION OF THE BIOLOGICALLY & STATISTICALLY SIGNIFICANT GENES

(i.e., significantly differentially expressed (DE) genes with False Discovery Rate (FDR) <0.05).
FDR < 0.05 limits false positives (i.e., genes that appear significant by chance, but are not).
This is done by filtering the DE genes by adjusted p-value. 

```{r}
#GENES UPREGULATED IN LumA SUBTYPE.

lumA_sig_genes <- lumA_basal %>%
      filter(adj.P.Val < 0.05 & abs(logFC) > 1)

# nrow(lumA_sig_genes) #654 significant genes
View(lumA_sig_genes)

#Upregulated genes in LumA subtype
lumA_upreg_genes <- lumA_sig_genes %>%
      filter(logFC > 1)

# nrow(lumA_upreg_genes) #337 upregulated genes in LumA subtype
# head(lumA_upreg_genes, 10) #top 10 lumA upregulated genes
# print(lumA_upreg_genes)
View(lumA_upreg_genes)

#Takeaway = FOXA1, MLPH, CA12, TBC1D9 & SLC44A4 are upregulated in luminal subtypes (both lumA & lumB) in comparison to other subtypes (Basal and Her2).

#save upregulated lumA genes as csv
# write.csv(lumA_upreg_genes, "lumA_upreg_genes.csv", row.names = TRUE)
# lumA_upreg_genes <- read.csv("lumA_upreg_genes.csv") #load matrix

```

Scientific Literature: 
      - BCL2, FOXA1 and MLPH are widely considered as "intrinsic" luminal markers in breast cancer - so the results of the volcano plot are consistent (Bernardo et al., 2013)(Metovic et al., 2022)(Hwang et al., 2018)(Eom et al., 2016)(Chen et al., 2015)(Hassan et al., 2025)
      - CA12 is considered as a luminal phenotype marker - so the results of the volcano plot are consistent (Porras et al., 2022).
      - TBC1D9 is a widely recognized modulator of tumorigenesis in luminal breast cancer cell lines (Kothari et al., 2021)
      - SLC44A4 drives immune evasion and metastasis in lumB tumors (Liu and Chen, 2026)
      - SERPINA3 associated with poorer prognosis in luminal tumors(Zhou et al., 2023)
      - GREB1 is an estrogen-upregulated gene and is required for ER+ breast cancer cell growth, and its overexpression correlates to increased colony formation (Rae et al., 2005)(Mohammed et al., 2021)(Ghosh et al., 2000)(Fagan and Yee, 2008)(Cheng et al., 2018)

Citations: 
      - Bernardo, G. M., Bebek, G., Ginther, C. L., Sizemore, S. T., Lozada, K. L., Miedler, J. D., Anderson, L. A., Godwin, A. K., Abdul-Karim, F. W., Slamon, D. J., & Keri, R. A. (2013). FOXA1 represses the molecular phenotype of basal breast cancer cells. Oncogene, 32(5), 554–563. https://doi.org/10.1038/onc.2012.62
      - Metovic, J., Borella, F., D'Alonzo, M., Biglia, N., Mangherini, L., Tampieri, C., Bertero, L., Cassoni, P., & Castellano, I. (2022). FOXA1 in Breast Cancer: A Luminal Marker with Promising Prognostic and Predictive Impact. Cancers, 14(19), 4699. https://doi.org/10.3390/cancers14194699
      - Porras, L., Gorse, F., Thiombane, N. K., Gaboury, L., & Mader, S. (2022). CAXII Is a Surrogate Marker for Luminal Breast Tumors Regulated by ER and GATA3. Cancers, 14(21), 5453. https://doi.org/10.3390/cancers14215453
      - Kothari, C., Clemenceau, A., Ouellette, G., Ennour-Idrissi, K., Michaud, A., C-Gaudreault, R., Diorio, C., & Durocher, F. (2021). TBC1D9: An Important Modulator of Tumorigenesis in Breast Cancer. Cancers, 13(14), 3557. https://doi.org/10.3390/cancers13143557
      - Liu Z, Chen R. SLC12A8 Drives Immune Evasion and Metastasis in Luminal B Breast Cancer by Inducing CD8+ T-Cell Exhaustion via the TLR Signaling Pathway. Cancer Med. 2026 Mar;15(3):e71712. doi: 10.1002/cam4.71712. PMID: 41795804; PMCID: PMC12967574.
      - Hwang, K. T., Kim, K., Chang, J. H., Oh, S., Kim, Y. A., Lee, J. Y., Jung, S. H., & Choi, I. S. (2018). BCL2 Regulation according to Molecular Subtype of Breast Cancer by Analysis of The Cancer Genome Atlas Database. Cancer research and treatment, 50(3), 658–669. https://doi.org/10.4143/crt.2017.134 
      - Eom, Y. H., Kim, H. S., Lee, A., Song, B. J., & Chae, B. J. (2016). BCL2 as a Subtype-Specific Prognostic Marker for Breast Cancer. Journal of breast cancer, 19(3), 252–260. https://doi.org/10.4048/jbc.2016.19.3.252
      - Chen, L. Y., Tsang, J. Y., Ni, Y. B., Chan, S. K., Chan, K. F., Zhang, S., & Tse, G. M. (2015). Bcl2 and Ki67 refine prognostication in luminal breast cancers. Breast cancer research and treatment, 149(3), 631–643. https://doi.org/10.1007/s10549-015-3288-4
      - Hassan, W. A., El-Assmy, M., ElBanna, A. K., Harbieh, I., Noufal, N., Lotfy, H., Shemais, T. A. H., Haikal, O. A., Saber, M. M., & Ali, R. I. (2025). Associations between BCL-2 expression and different histopathological prognostic factors in different molecular subtypes of invasive breast carcinoma of no special type. Histology and histopathology, 40(6), 825–834. https://doi.org/10.14670/HH-18-831
      - Zhou, J., Zhu, M., Wang, Q., Deng, Y., Liu, N., Liu, Y., & Liu, Q. (2023). SERPINA3-ANKRD11-HDAC3 pathway induced aromatase inhibitor resistance in breast cancer can be reversed by HDAC3 inhibition. Communications biology, 6(1), 695. https://doi.org/10.1038/s42003-023-05065-w
      - Mohammed, H., D'Santos, C., Serandour, A. A., Ali, H. R., Brown, G. D., Atkins, A., Rueda, O. M., Holmes, K. A., Theodorou, V., Robinson, J. L., Zwart, W., Saadi, A., Ross-Innes, C. S., Chin, S. F., Menon, S., Stingl, J., Palmieri, C., Caldas, C., & Carroll, J. S. (2013). Endogenous purification reveals GREB1 as a key estrogen receptor regulatory factor. Cell reports, 3(2), 342–349. https://doi.org/10.1016/j.celrep.2013.01.010
      - Ghosh, M. G., Thompson, D. A., & Weigel, R. J. (2000). PDZK1 and GREB1 are estrogen-regulated genes expressed in hormone-responsive breast cancer. Cancer research, 60(22), 6367–6375. 
      - Fagan, D. H., & Yee, D. (2008). Crosstalk between IGF1R and estrogen receptor signaling in breast cancer. Journal of mammary gland biology and neoplasia, 13(4), 423–429. https://doi.org/10.1007/s10911-008-9098-0
      - Cheng, M., Michalski, S., & Kommagani, R. (2018). Role for Growth Regulation by Estrogen in Breast Cancer 1 (GREB1) in Hormone-Dependent Cancers. International journal of molecular sciences, 19(9), 2543. https://doi.org/10.3390/ijms19092543
      - Rae, J. M., Johnson, M. D., Scheys, J. O., Cordero, K. E., Larios, J. M., & Lippman, M. E. (2005). GREB 1 is a critical regulator of hormone dependent breast cancer growth. Breast cancer research and treatment, 92(2), 141–149. https://doi.org/10.1007/s10549-005-1483-4

```{r}
#GENES UPREGULATED IN LumB SUBTYPE.

#differential expression (DE) results for lumB vs. Basal subtypes
lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf) %>%
      rownames_to_column(var == "gene_name")

#filtering for significant DE genes
lumB_sig_genes <- lumB_basal %>%
      filter(adj.P.Val < 0.05 & abs(logFC) > 1)

lumB_upreg_genes <- lumB_sig_genes %>% filter(logFC >1)

# nrow(lumB_upreg_genes) #248 upregulated genes in lumB subtype
head(lumB_upreg_genes, 10) #top 10 lumB upregulated genes
# print(lumB_upreg_genes)
View(lumB_upreg_genes)

#Takeaway = FOXA1, MLPH, CA12, TBC1D9 & SLC44A4 are upregulated in luminal subtypes (both lumA & lumB) in comparison to other subtypes (Basal and Her2).

#save upregulated lumB genes as csv
# write.csv(lumB_upreg_genes, "lumB_upreg_genes.csv", row.names = TRUE)
# lumB_upreg_genes <- read.csv("lumB_upreg_genes.csv") #load matrix

```

```{r}
##GENES UPREGULATED IN BASAL SUBTYPE.

basal_upreg_genes <- lumA_basal %>%
      filter(logFC < -1)

# lumA_sig_genes <- lumA_basal %>%
#       filter(adj.P.Val < 0.05 & abs(logFC) > 1)

# basal_upreg_genes <- lumA_sig_genes %>%
#       filter(logFC < -1)

# nrow(basal_upreg_genes) #317 upregulated genes in Basal subtype
# head(basal_upreg_genes, 10) #top 10 basal upregulated genes
# print(basal_upreg_genes)
View(basal_upreg_genes)

#Takeaway = ROPN1, PPP1R14C, HORMAD1 & UGT8 are upregulated in Basal tumors in comparison to luminal subtypes

#save upregulated basal genes as csv
# write.csv(basal_upreg_genes, "basal_upreg_genes_matrix.csv", row.names = TRUE)
# basal_upreg_genes <- read.csv("basal_upreg_genes_matrix.csv") #load matrix

```

Scientific Literature: 
      - ROPN1 and ROPN1B is a known prognostic and predictive biomarker in aggressive basal-like and triple-negative breast cancer (Cintra et al., 2025)(Liu et al., 2020)
      - KRT16 is overexpressed in basal-like breast cancer (BLBC) and is associated with tumor aggressiveness. Also contributes to poor prognosis and metastasis (Elazezy et al., 2021). 
      - PPP1R14C is also highly expressed in basal-like breast cancer (BLBC) patients (Jian et al., 2022)
      - HORMAD1 is overexpressed in basal-like breast cancer (BLBC) patients (Wang et al., 2018)
      - UGT8 is a known prognostic biomarker in basal-like breast cancer (BLBC) patients (Cao et al. ,2018)

Citations:
      - Cintra, R. C., Céspedes, A. G., Gamarra, O. M., Melgarejo, C. J., & de Bastos, D. R. (2025). ROPN1 Gene Expression as a Prognostic and Predictive Biomarker in Aggressive Breast Cancer: Clinical Implications and Survival Association. European journal of breast health, 21(4), 345–355. https://doi.org/10.4274/ejbh.galenos.2025.2025-6-2
      - Jian, Y., Kong, L., Xu, H., Shi, Y., Huang, X., Zhong, W., Huang, S., Li, Y., Shi, D., Xiao, Y., Yang, M., Li, S., Chen, X., Ouyang, Y., Hu, Y., Chen, X., Song, L., Ye, R., & Wei, W. (2022). Protein phosphatase 1 regulatory inhibitor subunit 14C promotes triple-negative breast cancer progression via sustaining inactive glycogen synthase kinase 3 beta. Clinical and translational medicine, 12(1), e725. https://doi.org/10.1002/ctm2.725
      - Wang, X., Tan, Y., Cao, X., Kim, J. A., Chen, T., Hu, Y., Wexler, M., & Wang, X. (2018). Epigenetic activation of HORMAD1 in basal-like breast cancer: role in Rucaparib sensitivity. Oncotarget, 9(53), 30115–30127. https://doi.org/10.18632/oncotarget.25728
      - Cao, Q., Chen, X., Wu, X., Liao, R., Huang, P., Tan, Y., Wang, L., Ren, G., Huang, J., & Dong, C. (2018). Inhibition of UGT8 suppresses basal-like breast cancer progression by attenuating sulfatide-αVβ5 axis. The Journal of experimental medicine, 215(6), 1679–1692. https://doi.org/10.1084/jem.20172048
      - Liu, Q., Huang, X., Li, Q., He, L., Li, S., Chen, X., Ouyang, Y., Wang, X., & Lin, C. (2020). Rhophilin-associated tail protein 1 promotes migration and metastasis in triple negative breast cancer via activation of RhoA. FASEB journal : official publication of the Federation of American Societies for Experimental Biology, 34(8), 9959–9971. https://doi.org/10.1096/fj.201903281R
      -  Elazezy, M., Schwentesius, S., Stegat, L., Wikman, H., Werner, S., Mansour, W. Y., Failla, A. V., Peine, S., Müller, V., Thiery, J. P., Ebrahimi Warkiani, M., Pantel, K., & Joosse, S. A. (2021). Emerging Insights into Keratin 16 Expression during Metastatic Progression of Breast Cancer. Cancers, 13(15), 3869. https://doi.org/10.3390/cancers13153869 

```{r}
#GENES UPREGULATED IN Her2 SUBTYPES.

lumA_her2_sig_genes <- lumA_Her2 %>%
      filter(adj.P.Val < 0.05 & abs(logFC) >1)

her2_upreg_genes <- lumA_her2_sig_genes %>%
      filter(logFC < -1)

# nrow(her2_upreg_genes) #114  upregulated genes in Her2 subtype
# head(her2_upreg_genes, 10) #top 10 her2 upregulated genes
# print(her2_upreg_genes)
View(her2_upreg_genes)

#Takeaway = GRB7, C17orf37, GSDML, RRM2, PSMD3 & PNMT are upregulated in Her2 tumors compared to lumA subtype

#save upregulated basal genes as csv
# write.csv(her2_upreg_genes, "her2_upreg_genes.csv", row.names = TRUE)
# her2_upreg_genes <- read.csv("her2_upreg_genes.csv") #load matrix

```

Scientific Literature: 
      - GSDML, PNMT, C17orf37, PSMD3 and GRB7 are known to be overexpressed in Her2-positive breast cancer and are mostly associated with poor clinical outcome (Katz et al., 2010)(Sahlber et al., 2012)(Evans et al., 2006)(Pradip et al., 2013)(Luoh et al., 2019)(Hergueta-Redondo et al., 2016)(Nadler et al., 2010)(Hergueta-Redondo et al., 2016)(Rubio et al., 2021)
      - RRM2 overexpression is associated with enhancement of HER2 signaling and oncogenic pathways, and promotes invasion, metastasis and aggressive tumor behaviour (Shah et al., 2015)(Abdel-Rahman et al., 2022)(Zhang et al., 2014)
      - PSMD3 regulates breast cancer by stabilizing HER2 from degradation (Fararjeh et al., 2019)
      - AKR1B10 enhances the latter stages of metastasis (i.e., colonization of distant organs and growth at secondary site)(van Weverwijk et al., 2019)(Cao et al., 2022)(Reddy et al., 2017)
      - CLCA2 downregulated in her2 breast tumor; its expression is known to inhibit migration and invasion (Purrington et al., 2020)(Walia et al., 2012)(Gruber and Pauli, 1999)

Citations: 
      - Sahlberg, K. K., Hongisto, V., Edgren, H., Mäkelä, R., Hellström, K., Due, E. U., Moen Vollan, H. K., Sahlberg, N., Wolf, M., Børresen-Dale, A. L., Perälä, M., & Kallioniemi, O. (2013). The HER2 amplicon includes several genes required for the growth and survival of HER2 positive breast cancer cells. Molecular oncology, 7(3), 392–401. https://doi.org/10.1016/j.molonc.2012.10.012
      - Evans, E. E., Henn, A. D., Jonason, A., Paris, M. J., Schiffhauer, L. M., Borrello, M. A., Smith, E. S., Sahasrabudhe, D. M., & Zauderer, M. (2006). C35 (C17orf37) is a novel tumor biomarker abundantly expressed in breast cancer. Molecular cancer therapeutics, 5(11), 2919–2930. https://doi.org/10.1158/1535-7163.MCT-06-0389
      - Katz, E., Dubois-Marshall, S., Sims, A. H., Faratian, D., Li, J., Smith, E. S., Quinn, J. A., Edward, M., Meehan, R. R., Evans, E. E., Langdon, S. P., & Harrison, D. J. (2010). A gene on the HER2 amplicon, C35, is an oncogene in breast cancer whose actions are prevented by inhibition of Syk. British journal of cancer, 103(3), 401–410. https://doi.org/10.1038/sj.bjc.6605763
      - Pradip, D., Bouzyk, M., Dey, N., & Leyland-Jones, B. (2013). Dissecting GRB7-mediated signals for proliferation and migration in HER2 overexpressing breast tumor cells: GTP-ase rules. American journal of cancer research, 3(2), 173–195. 
      - Luoh, S. W., Wagoner, W., Wang, X., Hu, Z., Lai, X., Chin, K., Sears, R., & Ramsey, E. (2019). GRB7 dependent proliferation of basal-like, HER-2 positive human breast cancer cell lines is mediated in part by HER-1 signaling. Molecular carcinogenesis, 58(5), 699–707. https://doi.org/10.1002/mc.22963
      - Hergueta-Redondo, M., Sarrio, D., Molina-Crespo, Á., Vicario, R., Bernadó-Morales, C., Martínez, L., Rojo-Sebastián, A., Serra-Musach, J., Mota, A., Martínez-Ramírez, Á., Castilla, M. Á., González-Martin, A., Pernas, S., Cano, A., Cortes, J., Nuciforo, P. G., Peg, V., Palacios, J., Pujana, M. Á., Arribas, J., … Moreno-Bueno, G. (2016). Gasdermin B expression predicts poor clinical outcome in HER2-positive breast cancer. Oncotarget, 7(35), 56295–56308. https://doi.org/10.18632/oncotarget.10787
      - Fararjeh, A. S., Chen, L. C., Ho, Y. S., Cheng, T. C., Liu, Y. R., Chang, H. L., Chang, H. W., Wu, C. H., & Tu, S. H. (2019). Proteasome 26S Subunit, non-ATPase 3 (PSMD3) Regulates Breast Cancer by Stabilizing HER2 from Degradation. Cancers, 11(4), 527. https://doi.org/10.3390/cancers11040527
      - Nadler, Y., González, A. M., Camp, R. L., Rimm, D. L., Kluger, H. M., & Kluger, Y. (2010). Growth factor receptor-bound protein-7 (Grb7) as a prognostic marker and therapeutic target in breast cancer. Annals of oncology : official journal of the European Society for Medical Oncology, 21(3), 466–473. https://doi.org/10.1093/annonc/mdp346
      - Hergueta-Redondo, M., Sarrio, D., Molina-Crespo, Á., Vicario, R., Bernadó-Morales, C., Martínez, L., Rojo-Sebastián, A., Serra-Musach, J., Mota, A., Martínez-Ramírez, Á., Castilla, M. Á., González-Martin, A., Pernas, S., Cano, A., Cortes, J., Nuciforo, P. G., Peg, V., Palacios, J., Pujana, M. Á., Arribas, J., … Moreno-Bueno, G. (2016). Gasdermin B expression predicts poor clinical outcome in HER2-positive breast cancer. Oncotarget, 7(35), 56295–56308. https://doi.org/10.18632/oncotarget.10787
      - Rubio, A. J., Bencomo-Alvarez, A. E., Young, J. E., Velazquez, V. V., Lara, J. J., Gonzalez, M. A., & Eiring, A. M. (2021). 26S Proteasome Non-ATPase Regulatory Subunits 1 (PSMD1) and 3 (PSMD3) as Putative Targets for Cancer Prognosis and Therapy. Cells, 10(9), 2390. https://doi.org/10.3390/cells10092390
      - Shah, K. N., Wilson, E. A., Malla, R., Elford, H. L., & Faridi, J. S. (2015). Targeting Ribonucleotide Reductase M2 and NF-κB Activation with Didox to Circumvent Tamoxifen Resistance in Breast Cancer. Molecular cancer therapeutics, 14(11), 2411–2421. https://doi.org/10.1158/1535-7163.MCT-14-0689
      - Abdel-Rahman, M. A., Mahfouz, M., & Habashy, H. O. (2022). RRM2 expression in different molecular subtypes of breast cancer and its prognostic significance. Diagnostic pathology, 17(1), 1. https://doi.org/10.1186/s13000-021-01174-4
      - Zhang, H., Liu, X., Warden, C. D., Huang, Y., Loera, S., Xue, L., Zhang, S., Chu, P., Zheng, S., & Yen, Y. (2014). Prognostic and therapeutic significance of ribonucleotide reductase small subunit M2 in estrogen-negative breast cancers. BMC cancer, 14, 664. https://doi.org/10.1186/1471-2407-14-664
      - van Weverwijk, A., Koundouros, N., Iravani, M., Ashenden, M., Gao, Q., Poulogiannis, G., Jungwirth, U., & Isacke, C. M. (2019). Metabolic adaptability in metastatic breast cancer by AKR1B10-dependent balancing of glycolysis and fatty acid oxidation. Nature communications, 10(1), 2698. https://doi.org/10.1038/s41467-019-10592-4
      - Cao, Z., Delfino, K., Tiwari, V., Wang, X., Hannan, A., Zaidi, F., McClintock, A., Robinson, K., Zhu, Y., Gao, J., Cao, D., & Rao, K. (2022). AKR1B10 as a Potential Novel Serum Biomarker for Breast Cancer: A Pilot Study. Frontiers in oncology, 12, 727505. https://doi.org/10.3389/fonc.2022.727505
      - Reddy, K. A., Kumar, P. U., Srinivasulu, M., Triveni, B., Sharada, K., Ismail, A., & Reddy, G. B. (2017). Overexpression and enhanced specific activity of aldoketo reductases (AKR1B1 & AKR1B10) in human breast cancers. Breast (Edinburgh, Scotland), 31, 137–143. https://doi.org/10.1016/j.breast.2016.11.003
      - Purrington, K. S., Knight, J., 3rd, Dyson, G., Ali-Fehmi, R., Schwartz, A. G., Boerner, J. L., & Bandyopadhyay, S. (2020). CLCA2 expression is associated with survival among African American women with triple negative breast cancer. PloS one, 15(4), e0231712. https://doi.org/10.1371/journal.pone.0231712
      - Gruber, A. D., & Pauli, B. U. (1999). Tumorigenicity of human breast cancer is associated with loss of the Ca2+-activated chloride channel CLCA2. Cancer research, 59(21), 5488–5491. 
      - Walia, V., Yu, Y., Cao, D., Sun, M., McLean, J. R., Hollier, B. G., Cheng, J., Mani, S. A., Rao, K., Premkumar, L., & Elble, R. C. (2012). Loss of breast epithelial marker hCLCA2 promotes epithelial-to-mesenchymal transition and indicates higher risk of metastasis. Oncogene, 31(17), 2237–2246. https://doi.org/10.1038/onc.2011.392

----------------------------
----------------------------
VOLCANO PLOTS TO VISUALIZE GENES THAT ARE UPREGULATED AND 
DOWNREGULATED ACROSS SUBTYPES.

```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)

#lumA v. Basal Volcano Plot
top_genes <- lumA_basal %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumA_vs_basal_volcano_plot <- ggplot(lumA_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(
            title = "LumA v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()
print(lumA_vs_basal_volcano_plot)

#save ggplot
# ggsave("lumA_vs_basal_volcano_plot.png", plot = lumA_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

```

         Significant
             ▲
             |
  Basal ◄----|----► LumA
             |
             ▼
      Not Significant

Interpretation of lumA vs. Basal Volcano Plot:
      - Right side (positive values) = genes upregulated in luminal tumors (lumA and LumB) = FOXA1, MLPH, CA12, TBC1D9, SLC44A4, DNALI1
      - Left side (negative values)= genes upregulated in Basal tumors = ROPN1, PPP1R14C, HORMAD1, UGT8
      - Top = statistically significant genes = ROPN1, PPP1R14C, HORMAD1, UGT8, FOXA1, MLPH, CA12, TBC1D9, SLC44A4, DNALI1
      - Bottom = genes that are NOT statistically significant = not labeled on plot.

```{r}
#lumA v. Her2 Volcano Plot

# library(ggplot2)
# library(ggrepel)
# library(dplyr)

#lumA_Her2 contrast
lumA_Her2 <- topTable(fit2, coef = "LumA_vs_Her2", number = Inf)

# #Add gene name to lumA_Her2 matrix
lumA_Her2 <- lumA_Her2 %>% rownames_to_column(var = "gene_name")

top_genes_lumA_her2 <- lumA_Her2 %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumA_vs_her2_volcano_plot <- ggplot(lumA_Her2, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumA_her2, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(
            title = "LumA v. Her2 Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()
print(lumA_vs_her2_volcano_plot)

# Save the plot
# ggsave("lumA_vs_her2_volcano_plot.png", plot = lumA_vs_her2_volcano_plot, width = 8, height = 6, dpi = 300)

```

Interpetation of lumA vs. Her2 Volcano Plot:
      - Right side (positive values) = genes upregulated in LumA = BCL2, SERPINA3, GREB1, MAPT
      - Left side (negative values)= genes upregulated in Her 2 tumors = GRB7, PNMT, RRM2, GSDML, C17orf37, RSMD3, PSMD3

```{r}
#lumB v. Basal Volcano Plot

#lumB_basal contrast
lumB_basal <- topTable(fit2, coef = "LumB_vs_Basal", number = Inf)

# #Add gene name to lumB_basal matrix
lumB_basal <- lumB_basal %>% rownames_to_column(var = "gene_name")

top_genes_lumB_basal <- lumB_basal %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumB_vs_basal_volcano_plot <- ggplot(lumB_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumB_basal, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(
            title = "LumB v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()
print(lumB_vs_basal_volcano_plot)

# Save the plot
# ggsave("lumB_vs_basal_volcano_plot.png", plot = lumB_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

```

Interpretation of LumB vs. Basal Volcano Plot:

      - Right side (positive values) = genes upregulated in LumB tumors = FOXA1, MLPH, CA12, TBC1D9, SLC44A4, LRFN2
      - Left side (negative values) = genes upregulated in Basal tumors = ROPN1, PPP1R14C, ROPN1B, KRT16

```{r}
#Her2 v. Basal Volcano Plot

#lumB_basal contrast
Her2_basal <- topTable(fit2, coef = "Her2_vs_Basal", number = Inf)

# #Add gene name to lumA_Her2 matrix
Her2_basal <- Her2_basal %>% rownames_to_column(var = "gene_name")

top_genes_Her2_basal <- Her2_basal %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

Her2_vs_basal_volcano_plot <- ggplot(Her2_basal, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_Her2_basal, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(
            title = "Her2 v. Basal Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()
print(Her2_vs_basal_volcano_plot)

#save ggplot
# ggsave("Her2_vs_basal_volcano_plot.png", plot = Her2_vs_basal_volcano_plot, width = 8, height = 6, dpi = 300)

```

Interpretation of Her2 vs. Basal Volcano Plot:
      - Right side (positive values) = genes upregulated in Her2 tumors = FOXA1, MLP, LRFN2, SLC44A4, C17orf37, GPR160
      - Left side (positive values) = genes upregulated in Basal tumors = ROPN1, ROPN1B, HORMAD1, UGT8

```{r}
#lumB v. Her2 Volcano Plot

#double check this graph - prints the same as lumA vs Basal

#lumB_Her2 contrast
lumB_Her2 <- topTable(fit2, coef = "LumB_vs_Her2", number = Inf)

# #Add gene name to lumA_Her2 matrix
lumB_Her2 <- lumB_Her2 %>% rownames_to_column(var = "gene_name")

top_genes_lumB_Her2 <- lumB_Her2 %>% #subsets the most significant genes to label
      arrange(adj.P.Val) %>% #sorts by adjusted p-value in ascending order
      slice(c(1:10, (n()-9):n())) # first 10 rows (1:10) = most significant genes. last 10 rows "n()-9):n())" = least significant genes

lumB_vs_Her2_volcano_plot <- ggplot(lumB_Her2, aes(x = logFC, y = -log10(adj.P.Val))) +
      geom_point(alpha = 0.5) +
      geom_text_repel(data = top_genes_lumB_Her2, aes(label = gene_name)) +
      geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "red") +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red") +
      theme_minimal() +
      labs(
            title = "lumB v. Her2 Volcano Plot",
            x = "Log2 Fold Change",
            y = "-Log10 Adjusted P-value"
      )

# quartz()
print(lumB_vs_Her2_volcano_plot)

#save ggplot
# ggsave("lumB_vs_Her2_volcano_plot.png", plot = lumB_vs_Her2_volcano_plot, width = 8, height = 6, dpi = 300)

```

Interpretation of LumB vs. Her2 Volcano Plot:
      - Right side (positive values) = genes upregulated in lumB tumors = C6orf211, GREB, PARD6B, PPM1J
      - Left side (negative values) = genes upregulated in Her2 tumors = GRB7, CLCA2, PNMT, C17orf37, ATP13A5, AKR1B10

----------------------------
----------------------------
HEATMAP OF GENE EXPRESSION PATTERNS/HEATMAP OF TOP VARIABLE GENES ACROSS PAM50 SUBTYPES.

```{r}
#install.packages("pheatmap")
library(pheatmap)
library(RColorBrewer)

gene_variances <- apply(lim_matrix, 1, var, na.rm = TRUE) #calculate the variance for each gene

top_var_genes <- names(sort(gene_variances, decreasing = TRUE))[1:100] #select the top 100 most variable genes

heatmap_data <- lim_matrix[top_var_genes, ] #subset the top 100 variable genes

heatmap_scaled <- t(scale(t(heatmap_data))) #scales the data by standardizing each gene (row) - enhances visualization

annotation_col <- data.frame(PAM50 = pam50) #add pam50 annotations
rownames(annotation_col) <- colnames(heatmap_scaled)

#plot & save the heatmap
# png("Heatmap_of_Top_100_Variable_Genes.png", width = 800, height = 600, res = 150)
heatmap_plot <- pheatmap(
                        heatmap_scaled,
                        annotation_col = annotation_col,
                        show_rownames = FALSE,
                        show_colnames = FALSE,
                        clustering_method = "complete",
                        scale = "none",
                        main = "Heatmap of Top 100 Variable Genes Across PAM50 Subtypes"
                  )
# dev.off()

print(heatmap_plot)

# Takeaway = clear subtype-specific clustering (i.e., variable genes are clustered across distinct breast cancer subtypes).
# Clusters with higher expression = yellow/red.
# Clusters with lower expression = blue.
# Basal subtype (green) is seemingly more tightly grouped together (suggesting distinct gene expression).
# LumA and LumB appear to be overlapping/intermixed in some regions = suggesting biological similarity.
# Similar expression patterns across groups of genes suggest co-regulated gene sets and/or shared biological pathways. 


```

----------------------------
MERGE GENE EXPRESSION WITH SURVIVAL DATA

rows = patient
columns = gene expression + survival data

```{r}
#merge rna seq data and survival data

gene_expr <- as.data.frame(t(RNA_seq_data))
colnames(gene_expr) <- RNA_seq_data$gene_name #preseve gene names 
gene_expr$patient_barcode <- rownames(gene_expr)

survival_dataset <- dplyr::left_join(
      final_aggregated_patient_data,
      gene_expr,
      by = "patient_barcode"
)

# View(as.data.frame(survival_dataset))
View(survival_dataset)
# "overall_survival" %in% colnames(survival_dataset) #true
# "status" %in% colnames(survival_dataset) #true
# "demographic.vital_status" %in% colnames(survival_dataset) #true
# "demographic.days_to_death" %in% colnames(survival_dataset) #true

#save upregulated survival_dataset as csv
# write.csv(survival_dataset, "survival_dataset.csv", row.names = TRUE)
# survival_dataset <- read.csv("survival_dataset.csv") #load matrix

```

CREATE COLUMNS FOR SURVIVAL ANALYSIS

      - Censoring Status = values in this column are either:
            - 1 (i.e., Event Happened (i.e., Death)) #151
            - 0 (i.e., Censored (i.e., Alive)) #843
      - Time-To-Event = from the start time (i.e., time of diagnosis), to the end time (i.e., time when event (death) occured or censoring (i.e., last follow-up))

```{r}
library(dplyr)

survival_dataset <- survival_dataset %>%
      mutate(
            follow_up_days = as.numeric(diagnoses.days_to_last_follow_up), #for patients who are still alive/censored
            death_days = as.numeric(demographic.days_to_death), #for patients who have died
            time_to_event = ifelse(!is.na(death_days), death_days, follow_up_days),
            event_status = as.numeric(status)
      ) %>%
      filter(!is.na(time_to_event)) #filters patients with no follow-up time or no death-time

# table(survival_dataset$event_status)
table(survival_dataset$event_status, useNA = "always") # 0(alive) = 843, 1(deaths) = 151, NAs = 42
summary(survival_dataset$time_to_event)

# > table(survival_dataset$status, useNA = "always")
#    0    1 <NA> 
#  843  151   42 

#remove 42NAs from event_status
survival_dataset <- survival_dataset %>%
      filter(!is.na(event_status))

View(survival_dataset)

# > summary(survival_dataset$time_to_event)
#    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
#    -7.0   436.5   808.5  1207.1  1648.2  8556.0 

```

SURVIVAL OBJECT

```{r}
library(survival)

surv_obj <- Surv(
      time = survival_dataset$time_to_event,
      event = survival_dataset$event_status
)

View(surv_obj) # 0(alive) = 843, 1(deaths) = 151, NAs = 42

```

BUILD COX MODEL

```{r}
library(survival)
library(dplyr)

#Cox Model for BCL2 only
# ----------------------------

#extract the BCL2 column from the survival dataset
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, BCL2) %>%
      na.omit()

cox_data$BCL2 <- as.numeric(trimws(cox_data$BCL2)) #ensures that the BCL2 column is numeric.

#run cox model on BCL2 only
cox_model <- coxph(
      Surv(time_to_event, event_status) ~ BCL2,
      data = cox_data
)

summary(cox_model)

# Calculate hazard_ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

```

Results & Interpretation:
      - Concordance= 0.565  (se = 0.04 ) -> very weak predictive power
      - Likelihood ratio test= 0  on 1 df,   p=1 -> no evidence that BCL2 alone affects survival
      - Wald test            = 0  on 1 df,   p=1 
      - Score (logrank) test = 0  on 1 df,   p=1
      - BCL2 alone is not prognostic in my cohort (i.e., it does not influence overall survival in this breast cancer dataset)

Concordance Index (C-index) (validation of survival outcome) 
      - Indicates the model's predictive ability
      - Higher C-index = better prognosis. 

| C-index | Interpretation              |
| ------- | --------------------------- |
| 0.5     | No better than chance       |
| 0.6–0.7 | Poor to fair discrimination |
| 0.7–0.8 | Good discrimination         |
| 0.8–0.9 | Very good discrimination    |
| >0.9    | Excellent discrimination    |

Hazard Ratio:
     BCL2 
0.9973889 

- Hazard Ratio (HR) < 1 = higher expression decreases risk (i.e., BCL2 is a protective gene in this cohort)

```{r}
library(survival)
library(dplyr)

#cox model on 6 genes upreg in Basal subtype (from limma results)

#genes upreg in basal subtype
up_genes_basal <- c("ROPN1", "UGT8", "PPP1R14C", "HORMAD1", "KRT16", "ROPN1B")

#subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_basal)) %>%
      na.omit() #remove NA's

#make the gene columns numeric
cox_data[up_genes_basal] <- lapply(cox_data[up_genes_basal], function(x) as.numeric(trimws(x)))

#run multivariate cox model on said 6 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_basal, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate hazard_ratios
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

```

Results & Interpretation
      - Concordance= 0.548  (se = 0.037) -> weak predictive ability.
      - Likelihood ratio test= 7.42  on 6 df,   p=0.3 
      - Wald test            = 7.18  on 6 df,   p=0.3
      - Score (logrank) test = 7.24  on 6 df,   p=0.3
      - These 6 genes upregulated in the Basal subtype are not statistically significant and do not collectively predict survival in this cohort.
      - p = 0.3 -> rejects null hypothesis

Hazard Ratio:
    ROPN1      UGT8  PPP1R14C   HORMAD1     KRT16    ROPN1B 
0.8673444 1.1919325 0.9482355 0.9731184 1.1055914 0.9708975 

- Hazard Ratio (HR) > 1 = higher expression increases risk = worse survival (i.e., risk genes) - UGT8, KRT16
- Hazard Ratio (HR) < 1 = higher expression decreases risk (i.e., protective genes) - ROPN1, PPP1R14C, HORMAD1, ROPN1B

```{r}
library(survival)
library(dplyr)

#cox model on  genes upreg in luminal subtype (both lumA & lumB) (from limma results)

#genes upreg in luminal subtype
up_genes_luminal <- c("BCL2", "SERPINA3", "GREB1", "MAPT", "FOXA1", "CA12", "TBC1D9", "MLPH", "SLC44A4", "PARD6B", "PPM1J", "LRFN2")

#subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_luminal)) %>%
      na.omit() #remove NA's

#make the gene columns numeric
cox_data[up_genes_luminal] <- lapply(cox_data[up_genes_luminal], function(x) as.numeric(trimws(x)))

#run multivariate cox model on said 6 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_luminal, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate hazard_ratios
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

```

Results & Interpretation
      - Concordance= 0.641  (se = 0.04 ) -> moderate predictive ability, estimate is reasonably precise.
      - Likelihood ratio test= 27.35  on 12 df,   p=0.007 -> p < 0.05 = At least a few of the aforementioned 12 genes are significantly associated with survival
      - Wald test            = 28.02  on 12 df,   p=0.006 -> p < 0.05 = at least one gene contributes significantly to predicting survival
      - Score (logrank) test = 28.23  on 12 df,   p=0.005 -> p < 0.05 = signifies that this group of genes are prognostic.
      - p < 0.05 = model is statistically significant

Hazard Ratios:
     BCL2  SERPINA3     GREB1      MAPT     FOXA1      CA12    TBC1D9      MLPH      SLC44A4    PARD6B     PPM1J     LRFN2 
0.9370675 0.8810642 0.7394603 0.7714521 1.1116027 1.0078950 1.1428458 1.0839761    0.8746222  1.2564292 1.0171508 1.1336509 

- Hazard Ratio (HR) > 1 = higher expression increases risk = worse survival (i.e., risk genes) - FOXA1, CA12, TBC1D9, MLPH, PARD6B, PPM1J, LRFN2
- Hazard Ratio (HR) < 1 = higher expression decreases risk (i.e., protective genes) - BCL2, SERPINA3, GREB1, MAPT, SLC44A4

```{r}
library(survival)
library(dplyr)

#cox model on  genes upreg in Her2 subtype (from limma results)

#genes upreg in Her2 subtype
up_genes_her2 <- c("GRB7", "PNMT", "RRM2", "GSDML", "C17orf37", "CLCA2", "ATP13A5", "AKR1B10", "MLPH", "GPR160", "LRFN2")

#subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(up_genes_her2)) %>%
      na.omit() #remove NA's

#make the gene columns numeric
cox_data[up_genes_her2] <- lapply(cox_data[up_genes_her2], function(x) as.numeric(trimws(x)))

#run multivariate cox model on said 6 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(up_genes_her2, collapse = " + "))),
      data = cox_data
)

summary(cox_model)
# summary(cox_model)$coefficients

# Calculate hazard_ratios
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)
```

Results & Interpretation:
      - Concordance= 0.604  (se = 0.04 ) -> low-to-moderate predictive ability.
      - Likelihood ratio test= 16.24  on 11 df,   p=0.1 -> not statistically significant (p > 0.05)
      - Wald test            = 16.28  on 11 df,   p=0.1
      - Score (logrank) test = 16.43  on 11 df,   p=0.1 -> genes are not strongly associated with survival.
      - Collectively, this group of genes does not significantly predict survival.

Hazard Ratios:
     GRB7      PNMT      RRM2     GSDML  C17orf37     CLCA2   ATP13A5   AKR1B10     MLPH    GPR160     LRFN2   
1.2068815 1.0620634 1.0971158 1.1839978 0.7062145 0.8716347 0.9819922 1.0044451 0.9108290 1.3957046 1.0744616 
    
- Hazard Ratio (HR) > 1 = higher expression increases risk = worse survival (i.e., risk genes) - GRB7, PNMT, RRM2, GSDML, AKR1B10, GPR160, LRFN2 
- Hazard Ratio (HR) < 1 = higher expression decreases risk (i.e., protective genes) - C17orf37, CLCA2, ATP13A5, MLPH

```{r}
library(survival)
library(dplyr)

# Cox Model for Multi-gene Signature 
#DEGs from limma
genes_of_interest <- c("BCL2", "FOXA1", "UGT8", "ROPN1", "HORMAD1", "CA12", "SERPINA3", "GREB1", "MAPT", "TBC1D9", "SLC44A4", "LRFN2", "GSDML", "C17orf37", "PNMT", "RRM2", "GRB7")

#subset relevant columns
cox_data <- survival_dataset %>%
      select(time_to_event, event_status, all_of(genes_of_interest)) %>%
      na.omit() #remove NA's

#make the gene columns numeric
cox_data[genes_of_interest] <- lapply(cox_data[genes_of_interest], function(x) as.numeric(trimws(x)))

#run multivariate cox model on said 6 genes
cox_model <- coxph(
      as.formula(paste("Surv(time_to_event, event_status) ~", paste(genes_of_interest, collapse = " + "))),
      data = cox_data
)

summary(cox_model)

# Calculate hazard_ratio
hazard_ratio <- exp(coef(cox_model))
print(hazard_ratio)

```

Results & Interpretation:
      - Concordance index (c-index)= 0.655  (se = 0.04 ) -> moderate predictive ability.
      - Likelihood ratio test= 28.52  on 17 df,   p=0.04 -> statistically significant.
      - Wald test            = 29.58  on 17 df,   p=0.03
      - Score (logrank) test = 30.41  on 17 df,   p=0.02 -> Collectively, this group of genes is associated with survival.

Hazard Ratios:
     BCL2     FOXA1      UGT8     ROPN1   HORMAD1      CA12  SERPINA3     GREB1 
1.0086742 1.0392526 1.0940709 0.8729866 0.8990500 1.0236287 0.9345558 0.7742825 
     MAPT    TBC1D9   SLC44A4     LRFN2     GSDML  C17orf37      PNMT      RRM2 
0.7467375 1.1678291 0.8510021 1.1256147 0.9765185 0.9605693 0.9382773 0.9746458 
     GRB7 
1.0612451 

- Hazard Ratio (HR) > 1 = higher expression increases risk = worse survival (i.e., risk genes) = BCL2, FOXA1, UGT8, CA12, TBC1D9, LRFN2, GRB7
- Hazard Ratio (HR) < 1 = higher expression decreases risk (i.e., protective genes) = ROPN1, HORMAD1, SERPINA3, GREB1, MAPT, SLC44A4, GSDML, C17orf37, PNMT, RRM2 

----------------------------
----------------------------
CALCULATION OF RISK (PROGNOSTIC) SCORE (a.k.a MULTI-GENE SIGNATURE SCORE)

A risk score is calculated for each patient.
Risk scores are used to separate patients into HIGH RISK v. LOW RISK GROUPS.

It is this prognostic score that turns the model to into a predictive tool. 

```{r}
#add risk scores
cox_data$risk_score <- predict(cox_model, type = "lp")
# head(cox_data$risk_score)

positive_risk <- cox_data[cox_data$risk_score > 0, ]
nrow(positive_risk) #number of patients with positive risk scores = 227

negative_risk <- cox_data[cox_data$risk_score < 0, ]
nrow(negative_risk) #number of patients with negative risk scores = 241

#note, survival_dataset (994 patients). After filtering out 
#patients with missing values, the cox_data (used to fit the 
#cox model), now has 468 patients. None with missing values. 
# Cox model can only use patients with complete data. 

```

STRATIFY PATIENTS 
into "High Risk" v. "Low Risk" groups" using cox_data dataset

```{r}
#stratify patients by median risk score
median_score <- median(cox_data$risk_score)

cox_data <- cox_data %>%
      mutate(
            risk_group = ifelse(risk_score > median_score, "High Risk", "Low Risk")
      )

table(cox_data$risk_group)
# High Risk  Low Risk 
#       234       234 
```

Interpretation:
      - High risk = worse survival
      - Low risk = better survival

----------------------------
KAPLAN-MEIER PLOT 

Visualization of the comparison between "High Risk" v. "Low Risk" groups.

```{r}
library(survival)
library(survminer)

#creation of survival object
surv_obj <- Surv(time = cox_data$time_to_event, event = cox_data$event_status)

km_fit <- survfit(surv_obj ~ risk_group, data = cox_data) #fit kaplan-meier (high risk vs. low risk)

km_fit <- survfit(Surv(time_to_event, event_status) ~ risk_group, data = cox_data) #fit kaplan survival curves

#plot kaplan
kaplan <- ggsurvplot(
      km_fit,
      data = cox_data,
      risk.table = TRUE, 
      pval = TRUE,
      conf.int = TRUE,
      palette = c("red", "green"), #distinguish low risk (green) vs. high risk (red) by colour
      xlab = "Time (days)",
      ylab = "Overall Survival Probability",
      title = "Kaplan-Meier Survival Curves by Risk Group",
      legend.title = "Risk Group",
      legend.labs = c("High Risk", "Low Risk") #label high and low risk groups
)

print(kaplan)

#save kaplan plot
ggsave(
      filename = "KM_survival_plot.png",
      plot = kaplan$plot,
      width = 8,
      height = 6,
      dpi = 300
)

```

Interpretation:
      - High risk = patients above median = worse survival
      - Low risk = patients below median = better survival
      - p = 0.00001 (< 0.05) = statistically significant

----------------------------

ESTIMATE X-YEAR SURVIVAL
(i.e., the probability of surviving after x number of years for patients in the high expression v. low expression groups)
```{r}
summary(km_fit, times = c(365, 1095, 1825)) #time in days

```

Call: survfit(formula = Surv(time_to_event, event_status) ~ risk_group, 
    data = cox_data)

                risk_group=High Risk 
 time n.risk n.event survival std.err lower 95% CI upper 95% CI
  365    197       6    0.973  0.0110        0.951        0.995
 1095     99      19    0.848  0.0286        0.794        0.906
 1825     56       8    0.759  0.0397        0.685        0.841

                risk_group=Low Risk 
 time n.risk n.event survival std.err lower 95% CI upper 95% CI
  365    211       3    0.986  0.0079        0.971        1.000
 1095    131       7    0.945  0.0171        0.912        0.979
 1825     86       6    0.891  0.0270        0.839        0.945

Interpretation:
      -  High Risk Group = Survival drops faster over time = more deaths over time.
            - 1-year survival = 97.3%
            - 3-year survival = 84.4%
            - 5-year survival = 75.9%
      -  Low Risk Group = Survival remains constant over time
            - 1-year survival = 98.6%
            - 3-year survival = 94.5%
            - 5-year survival = 89.1%

----------------------------

ESTIMATE MEDIAN SURVIVAL TIME
(i.e., the estimated time when survival = 0.5)

```{r}
summary(km_fit)

```

Interpretation:
      - Median survival time (High Risk Group) = 2,965 days (~8.1 years) = shorter survival
      - Median survival time (Low Risk Group) = 4,456 days (~12.2 years) = longer survival

----------------------------

UNIVARIATE COX REGRESSION ON RISK SCORE

Evaluates if multi-gene signature (i.e., risk score) actually predicts survival - and quantifies.

```{r}
library(survival)
library(dplyr)

coxph(Surv ~ risk_score)

cox_uni <- coxph(Surv(time_to_event, event_status) ~ risk_score,
      data = cox_data)

summary(cox_uni)

#run hazard ratio
exp(coef)

```

Results & Interpretation: 
      - Hazard Ratio (effect size) = 2.72 (for every 1-unit increase in risk score, theres a 2.7x increase in the hazard of death).
      - 95% Confidence Interval (CI) = 1.90 - 3.90.
      - p-value = 5.36e-08 = extremely significant predictor.
      - C-index = 0.655 = moderate predictive accuracy.

Univariate cox regression confirms that the risk score is 
significantly associated with overall survival (HR = 2.72, 95% CI = 1.90-3.90, p < 0.001).
Higher risk scores = increased hazard = multigene signatures is a strong predictor of patient prognosis.

----------------------------
----------------------------
KEY FINDINGS & BIOLOGICAL INSIGHTS/INTERPRETATION:

- Subtype-specific gene expression patterns and molecular signatures.
- Identified genes are consistent with known scientific literature.
- The model identified a gene signature associated with overall survival. 
      - Including GREB1, and MAPT (HR < 1). Others genes such as ROPN1, HORMAD1, SERPRINA3 showed a protective trend.
- Key genes that predicted poor prognosis in this cohort include UGT8 and TBC1D9 (HR > 1). Other genes such as FOXA1, BCL2, CA12 and GRB7 also showed a risk-increasing trend.
- C-index = 0.655 (se = 0.04) = moderate predictive ability.

Other key findings (supported by scientific literature): 
- GREB1 (HR < 1) is upregulated in low-risk patients and promote estrogen receptor signaling, luminal differentiation, and microtubule stability, consistent with their association with good prognosis.
- UGT8 (HR > 1) is upregulated in high-risk patients and promotes lipid metabolism, metastatic progression and cell survival, consistent with its association with poor prognosis and aggressive disease behaviour.
- BCL2 (HR > 1) is upregulated in high-risk patients and promote apoptosis inhibition, cell survival, and  therapy resistance consistent with its association with poor prognosis.

----------------------------
CONCLUSION & FUTURE WORK
----------------------------
- Gene expression differs across subtypes.
- Gene signatures can potentially be used to predict survival in breast cancer patients. 
- This analysis supports the potential of RNA-seq data for precision oncology.
- Future Work:
      - Validating the survival model on a different dataset.
      - Integrating more clinical variables (e.g., tumor stage and purity) to improve the model’s accuracy and predictive performance. 
      - Increasing sample size to improve predictive performance.
